---
title: "Econ 725: Machine Learning, Hw 1"
author: "Greg Kirwin"
date: "9/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,tidy.opts=list(width.cutoff=70),tidy=T) # allow text wrapping
library(data.table)
library(knitr)
library(formatR) # necessary to wrap comments and source code
library(ggplot2)
```

List of packages included[^1]: `data.table`, `ggplot2`, `stargazer`, `knitr`, `formatR`

[^1]: Adapted from previous solutions written by Dan McLeod (Thanks, Dan!)

## DB1B Data
The data require us to use BTS's download tool, so unfortunately we need to manually select the columns we want and then load in the data ourselves. Note: in previous semesters, we also asked students to scale up the passenger data by 10 since this is a 10% sample of all data. We did not do that this time, so the results are a little different from last year.

```{r db1b, echo = T}
#keep a copy of raw data
setwd("F:\\21fall\\725\\final project\\BTS data")

# read in as a data.table format, only shows example for 2014Q3
db1b_raw <- fread(file = "F:\\21fall\\725\\final project\\BTS data\\2014Q3_DB1B_MARKET.csv", header = T) 

#lower the column names for ease of use
colnames(db1b_raw) <- tolower(colnames(db1b_raw))

#start removing carrier changes, multiply passengers by 10, remove small markets and high fares
db1b <- db1b_raw[db1b_raw$tk_carrier_change == 0,]

#using data.table for the first time here to eliminate small markets
db1b = db1b[market_fare < 2500 & market_fare > 25,]
db1b <- db1b[,tp_mt := 1*(sum(passengers)<(365/4)*20),by=.(origin_airport_id,dest_airport_id)][tp_mt==0]

```

Next, we want to create a carrier-level and a market-level dataset.

````{r mkts, echo = T}
#Market-airline level

#need to weight the price and dist since passengers can be > 1
db1b_air = db1b[, .(airline_avgprice = weighted.mean(market_fare,passengers), 
                    airline_passangers = sum(passengers), 
                    airline_avgdist = weighted.mean(market_distance,passengers)), 
                by = .(origin_airport_id,dest_airport_id,ticket_carrier)]

#Market level
db1b_mkt = db1b[, .(mkt_avgprice = weighted.mean(market_fare, passengers),
                    mkt_avgdist = weighted.mean(market_distance,passengers),
                    mkt_numfirms = .N),
                by = .(origin_airport_id,dest_airport_id)]

#HHI: Need total passengers for each market, then get market shares, then square result
markets = db1b_air[, .(hhi = sum(((airline_passangers/sum(airline_passangers))*100)^2) ), by = .(origin_airport_id, dest_airport_id)]

#sort both (usually a good idea)
db1b_mkt = db1b_mkt[with(db1b_mkt, order(origin_airport_id, dest_airport_id)),]
markets = markets[with(markets, order(origin_airport_id, dest_airport_id)),]
db1b_mkt$mkt_hhi = markets$hhi

## Merge the market dataset with the populations.R data
#load(file="supp_files/populations.R")

#merged = merge(populations, db1b_mkt, by = c("origin_airport_id","dest_airport_id"))
````


### merge the airline level data and the market level data, only shows example for 2014Q3
````{r plots, echo = T}
merged = merge(db1b, db1b_mkt, by = c("origin_airport_id","dest_airport_id"))
merged14q3 = merge(merged, db1b_air, by = c("origin_airport_id","dest_airport_id","ticket_carrier"))
save(merged14q3,file = "merged14q3.r")

````



### only draw data of AA&US
````{r plots, echo = T}
AAUS12Q1 = merged12q1[ticket_carrier =="AA"| ticket_carrier=="US",]
save(AAUS12Q1,file = "AAUS12Q1.r")
AAUS12Q2 = merged12q2[ticket_carrier =="AA"| ticket_carrier=="US",]
save(AAUS12Q2,file = "AAUS12Q2.r")
AAUS12Q3 = merged12q3[ticket_carrier =="AA"| ticket_carrier=="US",]
save(AAUS12Q3,file = "AAUS12Q3.r")
AAUS12Q4 = merged12q4[ticket_carrier =="AA"| ticket_carrier=="US",]
save(AAUS12Q4,file = "AAUS12Q4.r")
AAUS14Q1 = merged14q1[ticket_carrier =="AA"| ticket_carrier=="US",]
save(AAUS14Q1,file = "AAUS14Q1.r")
AAUS14Q2 = merged14q2[ticket_carrier =="AA"| ticket_carrier=="US",]
save(AAUS14Q2,file = "AAUS14Q2.r")
AAUS14Q3 = merged14q3[ticket_carrier =="AA"| ticket_carrier=="US",]
save(AAUS14Q3,file = "AAUS14Q3.r")
AAUS14Q4 = merged14q4[ticket_carrier =="AA"| ticket_carrier=="US",]
save(AAUS14Q4,file = "AAUS14Q4.r")
````
